---
title: "Unifying spatially-explicit Rao's entropy, spatial concentration, and accumulation curves: the different aspects of local biodiversity"
# Output formats
format:
  stylisharticle-pdf:
    keep-tex: true
    # Color example, see "Text color"
    header-includes:
      \definecolor{grey}{RGB}{191, 191, 191}
  stylisharticle-html:
    # Color example, see "Text color"
    css: colors.css
# Authors (https://quarto.org/docs/journals/authors.html#author-schema)
author:
  - name: Eric Marcon
    orcid: 0000-0002-5249-321X
    affiliations:
      - name: AgroParisTech
        id: 1
        city: Montpellier
        state: France
      - name: AMAP, CIRAD, CNRS, INRAE, IRD, Univ Montpellier
        id: 2
        city: Montpellier
        state: France
    email: eric.marcon@agroparistech.fr
    attributes:
      corresponding: true
    url: https://ericmarcon.github.io/
  - name: Elodie Allié
    orcid: 0000-0001-8807-7982
  - name: Florence Puech
    orcid: 0000-0002-5279-6878
    affiliations:
      - name: Université Paris-Saclay, INRAE, AgroParisTech, Paris-Saclay Applied Economics
        id: 3
        city: Palaiseau
        state: France
  - name: Raphaël Pélissier
    orcid: 0000-0003-4845-5090
    affiliations:
      - name: AMAP, CIRAD, CNRS, INRAE, IRD, Univ Montpellier
        id: 2
        city: Montpellier
        state: France
# Abstract and Keywords
abstract: |
  To be written.
keywords: [template, demo]
# Bibliography
reference-section-title: References
bibliography: references.bib
# Language
lang: en-GB
# Code options 
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: false
  # Show messages
  message: false
  # Show warnings
  warning: false
  # Cache code results
  cache: false
# Template specific
journalinfo: "Preprint"
archive: "DOI: xxx/xx"
keywordlabel: Keywords
corrauthorlabel: Corresponding Author
---

```{r}
#| label: DoNotModify
### Utilities for R. 
# Do not modify unless you don't use R: then, delete this chunk.
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(c("knitR", "formatR", "kableExtra"))

# Chunk font size hook: allows size='small' or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

```{r}
#| label: Options
### Customized R options for this document
# Delete this chunk if you don't use R

# Add necessary packages here
packages <- c(
  "ads",
  "divent",
  "tidyverse"
)
# Install them
install_packages(packages)

# knitr options (https://yihui.org/knitr/options/)
knitr::opts_chunk$set(
  # Code chunk automatic format if tidy is TRUE
  tidy = TRUE, 
  # Tidy code options: remove blank lines and cut lines after 50 characters
  tidy.opts = list(blank = FALSE, width.cutoff = 50),
  # Font size in PDF output
  size = "scriptsize", 
  # Select PDF figures in PDF output if PDF file exists beside PNG file
  knitr.graphics.auto_pdf = TRUE
)
# Text width of R functions output
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

# Introduction

# Methods

## Ripley's K

The framework of spatial point pattern analysis, as introduced by @Ripley1977, relies on the K-function, defined so that $\lambda K(r)$ is the expected number of neighbours located within a distance $r$ of an arbitrary point of the pattern with intensity $\lambda$.
Its classical unbiased estimator for a point pattern located within a finite sampling window is

$$\hat{K}(r) = 
  \frac{1}{\hat{\lambda}n} 
  \sum_{i=1}^{n}{\sum_{j=1}^{n}{
    \mathbf{1}( \|x_i - x_j\| \leq r) c\left(i, j,,r\right)
  }}$$ {#eq-Khat}

where $n$ is the number of points of the pattern, $\hat{\lambda}$ is the pattern intensity generally estimated as $n/A$, with $A$ the sampling window area.
The indicator function $\mathbf{1}( \|x_i - x_j\| \leq r)$ equals 1 when points $i$ and $j$ are less than $r$ apart, 0 else.
$c\left(i, j,,r\right)$ is correction of edge effects equal to 1 when the circle of radius $r$ fits into the sampling window.
When it does not, the isotropic correction proposed by @Ripley1977 applies to windows of any shape thanks to appropriate geometrical considerations [@Goreaud1999] to estimate the number of neighbors whith no bias.

A common extension of this univariate function is the intertype (or bivariate) K12-function proposed by @Lotwick1982.
Its estimator is:

$$\hat{K_{1,2}}(r) =
  \frac{1}{\hat{\lambda_2}n_1}
  \sum_{i=1}^{n_1}{\sum_{j=1}^{n_2}{
    \mathbf{1}( \|x_i - x_j\| \leq r) c\left(i, j,,r\right)
  }}$$ {#eq-K12hat}

While $K(r)$ is a multiscale function of second-order neighbourhood properties of a point pattern, which allows robust tests of statistical significance against the null hypothesis of a random distribution of the points within the sampling window, $K_{12}(r)$ allows characterizing the spatial interaction between the two types of points, which can be tested against various null hypotheses, but in particular against the random labelling hypothesis, which means that, conditionally to the point locations, types 1 and 2 are randomly distributed among points of the pattern, or against the population independence hypothesis, meaning that, conditionally to the species-specific spatial patterns, population of type 1 points is independent from population of type 2 points [@Goreaud2003].



# Application

## Simulated data

## Paracou

# Conclusion
