---
title: "Unifying spatially-explicit Rao’s entropy, spatial concentration, and accumulation curves: the different aspects of local biodiversity"
# Output formats
format:
  stylisharticle-pdf:
    keep-tex: true
    # Color example, see "Text color"
    header-includes:
      \definecolor{grey}{RGB}{191, 191, 191}
  stylisharticle-html:
    # Color example, see "Text color"
    css: colors.css
# Authors (https://quarto.org/docs/journals/authors.html#author-schema)
author:
  - name: John Doe
    affiliations:
      - name: Dummy Organization
        department: Scientific Department
        address: Somewhere
        city: City
        country: Country
        postal-code: 9999
    orcid: 0000-0000-0000-0000
    email: JD@example.org
    url: https://example.org/
  - name: His Friend
  - name: Athird One
    corresponding: true
    email: at1@x.com
    url: https://www.x.com
    affiliations:
      - name: Third Affiliation
# Abstract and Keywords
abstract: |
  This document is only a demo explaining how to use the Stylish Article template.
keywords: [template, demo]
# Bibliography
reference-section-title: References
bibliography: references.bib
# Language
lang: en-GB
otherlangs: [fr-FR, it]
# Code options 
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: false
  # Show messages
  message: false
  # Show warnings
  warning: false
  # Cache code results
  cache: false
# Template specific
journalinfo: "Publication reference"
archive: "DOI: xxx/xx"
keywordlabel: Keywords
corrauthorlabel: Corresponding Author
---

```{r}
#| label: DoNotModify
### Utilities for R. 
# Do not modify unless you don't use R: then, delete this chunk.
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(c("knitR", "formatR", "kableExtra"))

# Chunk font size hook: allows size='small' or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

```{r}
#| label: Options
### Customized R options for this document
# Delete this chunk if you don't use R

# Add necessary packages here
packages <- c(
  "ads",
  "divent",
  "tidyverse"
)
# Install them
install_packages(packages)

# knitr options (https://yihui.org/knitr/options/)
knitr::opts_chunk$set(
  # Code chunk automatic format if tidy is TRUE
  tidy = TRUE, 
  # Tidy code options: remove blank lines and cut lines after 50 characters
  tidy.opts = list(blank = FALSE, width.cutoff = 50),
  # Font size in PDF output
  size = "scriptsize", 
  # Select PDF figures in PDF output if PDF file exists beside PNG file
  knitr.graphics.auto_pdf = TRUE
)
# Text width of R functions output
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

# Introduction

# Methods

## Ripley's K

The framework of spatial point pattern analysis, as introduced by @Ripley1977, relies on the K-function, defined so that $\lambda K(r)$ is the expected number of neighbours located within a distance $r$ of an arbitrary point of the pattern with intensity $\lambda$.
Its classical unbiased estimator for a point pattern located within a finite sampling window is

$$\hat{K}(r) = 
  \frac{1}{\hat{\lambda}N} 
  \sum_{i=1}^{n}{\sum_{j=1}^{n}{
    \mathbf{1}( \|x_i - x_j\| \leq r) c\left(i, j,,r\right)
  }}$$ {#eq-Khat}

where $N$ is the number of points of the pattern, $\hat{\lambda}$ is the pattern intensity generally estimated as $N/A$, with $A$ the sampling window area. The indicator function  $\mathbf{1}( \|x_i - x_j\| \leq r)$ equals 1 when points $i$ and $j$ are less than $r$ apart, 0 else. $c\left(i, j,,r\right)$ is correction of edge effects equal to 1 when the circle of radius $r$ fits into the sampling window. When it does not, the isotropic correction proposed by @Ripley1977 applies to windows of any shape thanks to appropriate geometrical considerations [@Goreaud1999] to estimate the number of neighbor whith no bias.

A common extension of this univariate function is the intertype (or bivariate) K12-function proposed by @Lotwick1982.

# Application

## Simulated data

## Paracou

# Conclusion
