---
title: "Unifying spatially-explicit Rao's entropy, spatial concentration, and accumulation curves: the different aspects of local biodiversity"
# Output formats
format:
  stylisharticle-pdf:
    keep-tex: true
    # Color example, see "Text color"
    header-includes:
      \definecolor{grey}{RGB}{191, 191, 191}
  stylisharticle-html:
    # Color example, see "Text color"
    css: colors.css
# Authors (https://quarto.org/docs/journals/authors.html#author-schema)
author:
  - name: Eric Marcon
    orcid: 0000-0002-5249-321X
    affiliations:
      - name: AgroParisTech
        id: 1
        city: Montpellier
        state: France
      - name: AMAP, CIRAD, CNRS, INRAE, IRD, Univ Montpellier
        id: 2
        city: Montpellier
        state: France
    email: eric.marcon@agroparistech.fr
    attributes:
      corresponding: true
    url: https://ericmarcon.github.io/
  - name: Elodie Allié
    orcid: 0000-0001-8807-7982
  - name: Florence Puech
    orcid: 0000-0002-5279-6878
    affiliations:
      - name: Université Paris-Saclay, INRAE, AgroParisTech, Paris-Saclay Applied Economics
        id: 3
        city: Palaiseau
        state: France
  - name: Raphaël Pélissier
    orcid: 0000-0003-4845-5090
    affiliations:
      - name: AMAP, CIRAD, CNRS, INRAE, IRD, Univ Montpellier
        id: 2
        city: Montpellier
        state: France
# Abstract and Keywords
abstract: |
  To be written.
keywords: [template, demo]
# Bibliography
reference-section-title: References
bibliography: references.bib
# Language
lang: en-GB
# Code options 
# https://quarto.org/docs/computations/execution-options.html#output-options
execute:
  # show code chunk output
  include: true
  # Show the code in the output
  echo: false
  # Show messages
  message: false
  # Show warnings
  warning: false
  # Cache code results
  cache: false
# Template specific
journalinfo: "Preprint"
archive: "DOI: xxx/xx"
keywordlabel: Keywords
corrauthorlabel: Corresponding Author
---

```{r}
#| label: DoNotModify
### Utilities for R. 
# Do not modify unless you don't use R: then, delete this chunk.
# Installation of R packages if necessary
install_packages <- function(packages) {
  invisible(
    sapply(
      packages, 
      FUN = function(package) {
        if (!package %in% installed.packages()[, 1]) {
          install.packages(
            package, 
            repos = "https://cran.rstudio.com/", 
            quiet = TRUE
          )
        }
      }
    )
  )
}

# Basic packages
install_packages(c("knitR", "formatR", "kableExtra"))

# Chunk font size hook: allows size='small' or any valid Latex font size in chunk options
def.chunk.hook <- knitr::knit_hooks$get("chunk")
knitr::knit_hooks$set(
  chunk = function(x, options) {
    x <- def.chunk.hook(x, options)
    ifelse(
      options$size == "normalsize", 
      yes = x,
      no = paste0("\n \\", options$size, "\n\n", x, "\n\n \\normalsize")
    )
  }
)
```

```{r}
#| label: Options
### Customized R options for this document
# Delete this chunk if you don't use R

# Add necessary packages here
packages <- c(
  "ads",
  "divent",
  "tidyverse"
)
# Install them
install_packages(packages)

# knitr options (https://yihui.org/knitr/options/)
knitr::opts_chunk$set(
  # Code chunk automatic format if tidy is TRUE
  tidy = TRUE, 
  # Tidy code options: remove blank lines and cut lines after 50 characters
  tidy.opts = list(blank = FALSE, width.cutoff = 50),
  # Font size in PDF output
  size = "scriptsize", 
  # Select PDF figures in PDF output if PDF file exists beside PNG file
  knitr.graphics.auto_pdf = TRUE
)
# Text width of R functions output
options(width = 50)

# ggplot style
library("tidyverse")
theme_set(theme_bw())
theme_update(
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA)
)
knitr::opts_chunk$set(dev.args = list(bg = "transparent"))

# Random seed
set.seed(973)
```

# Introduction

Local biotic and abiotic interactions are for long recognized as fundamental processes that shape the structure of ecological communities.
In particular, the spatial pattern of plant communities may result from the combined effects of seed dispersal, inter-individual (i.e. intra- and inter-specific) competition/facilitation and habitat preferences [@Dale1999].
While each single process may either promote or demote species coexistence, the diversity of interactions in space and/or time is a key feature for the maintenance of biological diversity [@Clark2010; @Cadotte2012].
As a consequence, the ecological literature reveals increasing efforts in integrating approaches from phylogenetics, functional and community ecology [@Mouquet2012], in order to account for the key components of species interactions with their biotic and abiotic environments.

<!--# TODO: introduce ecological questions addressed in appliucation section -->

Towards this goal, a number of novel analytical methods incorporating measures of between-species phylogenetic relatedness or functional trait dissimilarities as a surrogate for ecological differences have been developed [@Pavoine2011].
Among these methods some are specifically designed to explicitly account for space [e.g. @Pavoine2011a].
However those studies generally consider spatially discontinuous community samples, i.e. the $\beta$-diversity component of @Whittaker1960, so that the underlying spatial framework can basically be related to the graph theory [@Dale2002].
When mapped data are concerned, such as for instance those on tropical trees provided by large permanent forest plots [@Losos2004], the well-developed framework of the spatial point pattern analysis [@Ripley1976a; @Ripley1977] offers an efficient means of studying the continuous spatial variation in species diversity [@Condit2000; @Shimatani2001; @Rajala2012].
A synthesis of the opportunities brought by this approach can be found in @Wiegand2017.

<!--# TODO: developments of spatially explicit diversity indepent from point processes: neighbourhood diversity. -->

@Shen2013 and @Pelissier2015 concomitantly introduced within the framework of the spatial point pattern analysis a multitype correlation function that provides a distance-dependent measure of species diversity accounting for non-uniform species differences.
If both contributions independently arrived to very similar functions, we believe that to help ecologists to well-understand and to take full advantage of novel approach, further clarifications are needed, especially regarding how it is related to well-known functions of spatial point pattern analysis, namely Ripley's K-function [@Ripley1977] and its intertype extension, the $K_{1,2}$-function [@Lotwick1982] on the one hand, and Rao's quadratic entropy on the other hand [@Rao1982].

The aim of this chapter is thus to introduce the rationale behind the functions of multivariate point pattern analysis that integrate between-species differences in a distance–dependent measure of diversity.
It can basically be seen as an extension of the work of @Shimatani2001 based on Simpson's entropy to Rao's entropy, one of the most powerful and flexible measure of species diversity [@Pavoine2012].
It is also a particular implementation of the measurement of neighbourhood diversity which possesses some specific features due to the quadratic form of Rao's entropy.
The methods section hereafter makes the synthesis between these originally independent developments and outlines the central role of Rao's entropy to reconcile them.
We finally provide worked examples based on simulated data and on a real dataset from an experimental tropical forest plot in French Guiana.

# Methods

## Biodiversity measurement based on information theory

### Definitions

<!--# TODO: recall entropy, diversity, order 2, phylo/functional -->

### Diversity accumulation curves

<!--# TODO: recall SAC, generalise to DAC and introduce EAC -->

## Distance-dependent entropy based on the point-process theory

### Ripley's K

The framework of spatial point pattern analysis, as introduced by @Ripley1977, relies on the K-function, defined so that $\lambda K(r)$ is the expected number of neighbours located within a distance $r$ of an arbitrary point of the pattern with intensity $\lambda$.
Its classical unbiased estimator for a point pattern located within a finite sampling window is

$$\hat{K}(r) = 
  \frac{1}{\hat{\lambda}n} 
  \sum_{i=1}^{n}{\sum_{j=1}^{n}{
    \mathbf{1}( \|x_i - x_j\| \leq r) c\left(i, j,,r\right)
  }}$$ {#eq-Khat}

where $n$ is the number of points of the pattern, $\hat{\lambda}$ is the pattern intensity generally estimated as $n/A$, with $A$ the sampling window area.
The indicator function $\mathbf{1}( \|x_i - x_j\| \leq r)$ equals 1 when points $i$ and $j$ are less than $r$ apart, 0 else.
$c\left(i, j,,r\right)$ is correction of edge effects equal to 1 when the circle of radius $r$ fits into the sampling window.
When it does not, the isotropic correction proposed by @Ripley1977 applies to windows of any shape thanks to appropriate geometrical considerations [@Goreaud1999] to estimate the number of neighbors whith no bias.

A common extension of this univariate function is the intertype (or bivariate) K12-function proposed by @Lotwick1982.
Its estimator is:

$$\hat{K}_{1,2}(r) =
  \frac{1}{\hat{\lambda_2}n_1}
  \sum_{i=1}^{n_1}{\sum_{j=1}^{n_2}{
    \mathbf{1}( \|x_i - x_j\| \leq r) c\left(i, j,,r\right)
  }}$$ {#eq-K12hat}

$K(r)$ is a multiscale function of second-order neighbourhood properties of a point pattern, which allows robust tests of statistical significance against the null hypothesis of a random distribution of the points within the sampling window.
In contrast, $K_{12}(r)$ allows characterizing the spatial interaction between the two types of points, which can be tested against various null hypotheses.
The most interesting ones are the random labelling hypothesis, which means that, conditionally to the point locations, types 1 and 2 are randomly distributed among points of the pattern, and the population independence hypothesis, meaning that, conditionally to the species-specific spatial patterns, population of type 1 points is independent from population of type 2 points [@Goreaud2003].

### A distance-dependent Simpson's entropy

Simpson's index of species diversity [@Simpson1949] gives the probability that a randomly selected pair of individuals in a sample is comprised of two different species.
It is more than an index: it is a measure of entropy [@Jost2006] and thus should be named Simpson's entropy.
It is estimated with no bias by:

$$\hat{D} = 1 - \sum_{s =1}^S{\frac{n_s \left(n_s - 1 \right)}{n \left(n - 1 \right)}}$$ {#eq-Simpson}

where $n$ and $S$ are the total numbers of individuals and species in the sample, respectively, and $n_p$ the number of individuals of species $s$.

From this entropy, @Shimatani2001 proposed a distance-dependent version based on Ripley's K-function, from which we can write a version standardized by $\hat{D}$ as

$$\hat{K}_S(r) = \left[ 1 - 
  \sum_{s =1}^S{\frac{\hat{\lambda}_s^2 \hat{K}_s(r)}{\hat{\lambda}^2 \hat{K}(r)}}
  \right] / \hat{D}
  $$ {#eq-Ks}

where $\hat{K}(r)$ and $\hat{K}_s(r)$ are Ripley's K-functions for all points of the pattern (no distinction of species) and for points of species s, respectively, with $\hat{\lambda}$ and $\hat{\lambda}_s$ the corresponding intensity estimators.

The numerator of the above equation corresponds to function $\hat{\alpha}(r)$ of @Shimatani2001, that we may name Simpson's neighbourhood entropy.
Neighbourhood entropy (or diversity) is the entropy (or diversity) considered within a circle of radius $r$ around each point.
The interest of considering local diversity rather than at the scale of the whole window has emerged in the literature since @Batty1974.
A review can be found in @Altieri2021, with diverse approaches.
It is not discussed that the functioning of a community is influenced by its biodiversity, but it is straightforward to consider that each individual is concerned by the diversity of its neighbourhood rather that of the whole community [e.g., @Hubbell2001a showed that tree mortality increased with neighbourhood species richness in a tropical evergreen forest].

$\alpha(r)$ tends to $D$ when $r$ increases so $K(r)$ tends to 1.
It is noticeable that $\hat{K}_S(r)$ has an expectation of 1 for all $r$ under the random labelling hypothesis, i.e. a random distribution of the species names conditionally to the overall spatial pattern [@Eckel2008].

Using the unbiased estimator of $\hat{K}(r)$ introduced in @eq-Khat, allows correcting for edge effects for sampling windows of any shape.

We can notice for the following that $\hat{K}_S(r)$ can also be estimated by considering the pairs of heterospecifics [@Shimatani2001, eq. 9]:

$$\hat{K}_S(r) = \left[
  \sum_{s=1}^S{\sum_{t=1, s\ne t}^S{\frac{\hat{\lambda}_s \hat{\lambda}_t \hat{K}_{s,t}(r)}{\hat{\lambda}^2 \hat{K}(r)}}}
  \right] / \hat{D}
  $$ {#eq-Ks12}

with $\hat{K}_{s,t}(r)$ a generalization of $K_{12}(r)$ estimator introduced earlier, to any couple of different species $s$ and $t$.

### A generalization to Rao's entropy

Simpson's entropy is a special case of Rao's entropy, that is the expected divergence between individuals of a randomly selected pair in a sample [@Rao1982].
An unbiased estimator [@Shimatani2001a] is:

$$\hat{H}_D = \sum_{s =1}^S{\frac{n_s n_t}{n \left(n - 1 \right)}d_{s,t}}$$

where and $d_{s,t}$ is a measure of the pairwise divergence between species $s$ and $t$ (with $d_{s,t}$ = $d_{t,s}$ and $d_{s,s} = 0$, but the triangular inequality $d_{s,t} + d_{s,u} \geq d_{s,u}$ that defines a distance is not required).
$H_D$ is a measure of diversity commonly used in ecology to take into account non-uniform distances between species, such as phylogenetic or functional distances [@Pavoine2011].
Taking $d_{s,t} = 1$ for all $s \neq t$ and 0 otherwise, $H_D$ reduces to Simpson's $D$ above.
We therefore straightforwardly propose to generalize $K_S(r)$ introduced above to a normalized, distance-dependent Rao's quadratic entropy $K_S(r)$ that integrates Ripley's correction of edge effects estimated by:

$$\hat{K}_R(r) = \left[
  \sum_{s=1}^S{\sum_{t=1, s\ne t}^S{\frac{\hat{\lambda}_s \hat{\lambda}_t \hat{K}_{s,t}(r) d_{s,t}}{\hat{\lambda}^2 \hat{K}(r)}}}
  \right] / \hat{H}_D
  $$

This function has an expectation of 1 under the random labelling hypothesis and gives the mean proportion of Rao's sample diversity (i.e., $H_D$) in a disc of radius r centred on an arbitrary point of the pattern.
$\hat{K}_R(r)$ reduces to $\hat{K}_S(r)$ when $d_{s,t} = 1$ for all species $s \neq t$.

### Multitype correlation and $\beta$ diversity

Our $K_R$- and $K_S$-functions are cumulative functions as is the original K-function of @Ripley1977.
This means that they are estimated for a series of concentric neighbourhood discs of increasing radii, for instance by a regular distance interval $dr$.
Some authors however prefer to refer to the framework of the pair-correlation function, $g(r)$ [@Stoyan1987; @Wiegand2004], that can be seen as the derivative of $K(r)$, giving the probably to encounter neighbouring points in concentric annuli between $r$ and $r - dr$.
$K_d(r)$ of @Shen2013 is such a function, very close to $g_R(r)$, the derivative of our $K_R(r)$ above [@Pelissier2015].
$K_d(r)$ is therefore an awkward appellation for Shen's et al. function, which would have been better named as $g_d(r)$ as to remain consistent with the classical notation used in point pattern analysis.

The same reasoning can be built from the so-called Simpson's $\beta$ diversity index to $g_R(r)$.
@Chave2002 defined the $F(r)$ function as the probability for two individuals located $r$ apart to belong to the same species.
@Shen2013 named $1 - F(r)$ the "spatially explicit Simpson's \beta index of diversity", denoted $\beta(r)$ by @Shimatani2001, that is the probability for two individuals located $r$ apart to belong different species.
We will rather write $\beta_S(r)$ for consistency with our notation system.
It can be straightforwardly be generalized to $\beta_R(r)$, the average divergence between to individuals taken at distance $r$, thus a spatially explicit Rao's \beta index of diversity.
These measures are actually $\beta$ differentiation diversity measures sensu @Jurasinski2009: they characterize how different spatial units $r$ apart are.

We can thus define $g_S(r)$, a normalized version of $\beta_S(r)$ and $g_R(r)$, its generalization to Rao's entropy.

<!--# TODO: add formulas, check consistency with @Pelissier2015 -->

## Synthesis: a consistent measure of neighbourhood diversity

<!--# TODO: summarize the story around Rao's entropy at a distance -->

## Testing ecological models

The $K_R$-function allows computing phylogenetic or functional entropy in the neighbourhood of individuals with respect to distance.
Beyond measuring diversity, it also allows inferring ecological processes responsible of the observed patterns by testing them against appropriate null hypotheses.

### The random labelling null hypothesis

The null hypothesis of random labelling originally proposed by @Shimatani2001 provides a test of departure of $\hat{K}_R(r)$ from a random distribution of the mark labels among points of the pattern.
It allows testing for an absence of interaction between neighbouring points, conditionally to the overall pattern of point locations, so that the probability to find a pair of identical marks is independent of distance.
When the function is above 1, the mean diversity within a disc of radius r from an arbitrary point of the pattern is higher than the total sample diversity, and conversely.
Local confidence limits of the random labelling hypothesis at a given significance level can be built from Monte Carlo rank tests @Diggle1979.
But this hypothesis basically relies on the assumption that the marks result from a random process that affects individuals a posteriori with regards to the process of point location, so that we have for all $s$ types of marks $K_s(r) = K(r)$ [@Goreaud2003; @Eckel2008] and hence $K_R(r) = 1$.
This could for instance corresponds to a multitype process resulting from a disease attack in a plant population that would mark the individuals as sick, non-infected or healthy carrier [see also @Eckel2008 for an example in socio-economics].
Sounded reasons for testing the random labelling hypothesis are however hardly conceivable in the context of species diversity, so that testing this hypothesis only permits limited inferences about the processes at the origin of the observed pattern of species diversity.

### The population independence null hypothesis

A more meaningful hypothesis for species diversity studies is the one of mutually independent populations, postulating that the spatial pattern of each type of mark (or population) is the realization of a specific spatial process (for instance of limited dispersion) and that the location of points of each population is independent from the location of the points of the other populations in the resulting multivariate pattern.
This is a generalization of the population independence hypothesis proposed for testing the bivariate $K_{1,2}$-function by @Lotwick1982 for which we expect that $K_{s,t}(r) = \pi r^2$ for any marks $s \neq t$ [@Eckel2008], and hence, here also, $K_R(r) = 1$.
Note that this hypothesis differs from the random labelling hypothesis, except when all specific spatial patterns result from a stationary Poisson point process [@Diggle1983].

A classical method for testing the mutually independent population hypothesis is to build a Monte Carlo rank test keeping the pattern of each population unchanged, but randomizing their relative positions by a random shifting procedure of each pattern around a torus [@Goreaud2003].
Such a procedure is cumbersome for highly multivariate patterns and is thus abandoned.

### The species equivalence null hypothesis

More interestingly $K_R(r)$ provides opportunities for testing a highly relevant ecological hypothesis for diversity studies, based on $d_{s,t}$, which quantifies how species differ one from each other.
Shuffling between-species distances a large number of times means eliminating their differences and therefore testing a null hypothesis of species equivalence, put forward, for instance, in the neutral theory of biodiversity [@Hubbell2001].
This null hypothesis can be tested conditionally to the pattern of point locations and to the spatial distribution of the species labels by permuting simultaneously the rows and columns in $d_{s,t}$, which is equivalent, in the case of a cophenetic distance matrix, to permuting the tips of a hierarchical (taxonomic or phylogenetic) tree [@Hardy2008a].
It is noteworthy that under the null hypothesis of species equivalence, $K_S(r)$ provides a theoretical expectation of $K_R(r)$, i.e. the value expected when $d_{s,t}$ is a uniform between-species distance matrix.
When focusing on the species equivalence hypothesis, $K_R(r) / K_S(r)$, the standardization proposed by @Shen2013, provides a constant expectation of 1 at all distances.
With reference to these authors and to the usual notation in point pattern analysis, this ratio is appropriately designated as $K_d(r)$.

# Application

## Simulated data

We provide here a few examples to illustrate how the $K_R$-function behaves.
In a first experiment, we simulated habitat preferences by generating heterogeneous spatial patterns in a checked 1-ha plot with two habitats: eight species were simulated with a twofold higher intensity in habitat 1, and 8 species with a twofold higher intensity in habitat 2, so that species abundant in one habitat were rare in the other one and vice versa <!--# (Fig. 1) -->.

## Paracou

# Conclusion
